---
title: "MP01"
---

```{r}
#| code-fold: true
#| code-summary: "Show the code"
```

## Task 1: Data Acquisition
# Downloading the files.
```{r}
if(!dir.exists(file.path("data", "mp01"))){
    dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE)
}

GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.csv")

if(!file.exists(GLOBAL_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv", 
                  destfile=GLOBAL_TOP_10_FILENAME)
}

COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.csv")

if(!file.exists(COUNTRY_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv", 
                  destfile=COUNTRY_TOP_10_FILENAME)
}
```

```{r}
if(!require("tidyverse")) install.packages("tidyverse")
library(readr)
library(dplyr)

GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME)
```

```{r}
str(GLOBAL_TOP_10)
```

```{r}
glimpse(GLOBAL_TOP_10)
```

## Task 2: Data Cleaning
# Fix "N/A" values in season_title to NA.
```{r}
GLOBAL_TOP_10 <- GLOBAL_TOP_10 |>
  mutate(season_title = if_else(season_title == "N/A", NA, season_title))

# Confirm the fix
glimpse(GLOBAL_TOP_10)
```
```{r}
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = "N/A")

# Confirm the fix
glimpse(COUNTRY_TOP_10)
```
```{r}
COUNTRY_TOP_10 <- COUNTRY_TOP_10 |>
  mutate(season_title = if_else(season_title == "N/A", NA, season_title))

# Confirm the fix
glimpse(COUNTRY_TOP_10)
```

## Task 3: Data Import

```{r}
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = "N/A")


# Check the structure
glimpse(COUNTRY_TOP_10)

```

```{r}
if(!require("DT")) install.packages("DT")
library(DT)
```

# Show only first 20 rows.
```{r}
library(DT)
GLOBAL_TOP_10 |> 
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE))
```

# Formatting columns.
```{r}
library(stringr)
format_titles <- function(df){
    colnames(df) <- str_replace_all(colnames(df), "_", " ") |> str_to_title()
    df
}

GLOBAL_TOP_10 |> 
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

# Drop season_title
```{r}
GLOBAL_TOP_10 |> 
    select(-season_title) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

# converting runtime: hours to minutes.
```{r}
GLOBAL_TOP_10 |> 
    mutate(`runtime_(minutes)` = round(60 * runtime)) |>
    select(-season_title, 
           -runtime) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))
```

## Task 4: Exploratory Questions

```{r}
# Required Packages
if(!require("tidyverse")) install.packages("tidyverse")
if(!require("DT")) install.packages("DT")
if(!require("lubridate")) install.packages("lubridate")

library(dplyr)
library(readr)
library(DT)
library(lubridate)

```

# Q1. How many different countries does Netflix operate in? 
# Netflix operates in 94 countries.

```{r}
country_count <- COUNTRY_TOP_10 |> 
  distinct(country_iso2) |> 
  nrow()
```

# Q2. Which non-English-language film has spent the most cumulative weeks in the global top 10? How many weeks did it spend?
# The non-English film with the longest global Top 10 run is All Quiet on the Western Front with 23 weeks.

```{r}
non_english_film <- GLOBAL_TOP_10 |>
  filter(category == "Films (Non-English)") |>
  group_by(show_title) |>
  summarise(total_weeks = max(cumulative_weeks_in_top_10, na.rm = TRUE)) |>
  arrange(desc(total_weeks)) |>
  slice(1)
```

# Q3. What is the longest film (English or non-English) to have ever appeared in the Netflix global Top 10? How long is it in minutes?
# The longest film is Pushpa 2: The Rule (Reloaded Version) at 224 minutes.

```{r}
longest_film <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films")) |>
  filter(!is.na(runtime)) |>
  mutate(runtime_minutes = round(60 * runtime)) |>
  arrange(desc(runtime_minutes)) |>
  slice(1)
```

# Q4. For each of the four categories, what program has the most total hours of global viewership?

```{r}
if(!require(scales)) install.packages("scales")
library(scales)
GLOBAL_TOP_10 |>
  mutate(show_title = if_else(is.na(season_title), show_title, season_title)) |>
  group_by(category, show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE), .groups = "drop") |>
  group_by(category) |>
  slice_max(order_by = total_hours, n = 1, with_ties = FALSE) |>
  arrange(category) |>
  mutate(total_hours = comma(total_hours)) |>
  datatable(
    options = list(pageLength = 5, searching = FALSE, info = FALSE),
    rownames = FALSE,
    caption = "Programs with the Most Total Global Hours by Category"
  )
```
# Q5. Which TV show had the longest run in a country’s Top 10? How long was this run and in what country did it occur?
# Money Heist had the longest run with 127 weeks in Pakistan.

```{r}
longest_country_run <- COUNTRY_TOP_10 |>
  filter(str_detect(category, "TV")) |>
  group_by(country_name, show_title) |>
  summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(max_weeks)) |>
  slice(1)
```

# Q6. Netflix provides over 200 weeks of service history for all but one country in our data set. Which country is this and when did Netflix cease operations in that country?
# Russia with 35 weeks; the last week in the dataset is February 27, 2022.

```{r}
country_coverage <- COUNTRY_TOP_10 |>
  group_by(country_name) |>
  summarise(
    weeks_count  = n_distinct(week),
    last_week = max(week, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(weeks_count < 200) |>
  arrange(weeks_count)
```

# Q7. What is the total viewership of the TV show Squid Game? 
# Squid game has total hours of 5,048,300,000 globally (all seasons combined.)

```{r}
squid_game_hours <- GLOBAL_TOP_10 |>
  filter(show_title == "Squid Game") |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE)) |>
  pull(total_hours)
```

# Q8. The movie Red Notice has a runtime of 1 hour and 58 minutes. Approximately how many views did it receive in 2021?
# In 2021, Red Notice received 0 views.

```{r}
red_notice_2021 <- GLOBAL_TOP_10 |>
  filter(show_title == "Red Notice", year(week) == 2021) |>
  summarise(total_views = sum(weekly_views, na.rm = TRUE)) |>
  pull(total_views)
```

# Q9. How many Films reached Number 1 in the US but did not originally debut there?
# 45 films reached #1 in the US. The most recent was "KPop Demon Hunters.

```{r}
us_films <- COUNTRY_TOP_10 |>
  filter(country_name == "United States", str_detect(category, "Films")) |>
  group_by(show_title) |>
  arrange(week) |>
  mutate(
    ever_hit_1 = any(weekly_rank == 1),
    debut_rank = first(weekly_rank)
  ) |>
  filter(ever_hit_1 == TRUE, debut_rank > 1) |>
  ungroup() |>
  distinct(show_title) |>
  nrow()

most_recent_film <- COUNTRY_TOP_10 |>
  filter(country_name == "United States", str_detect(category, "Films")) |>
  group_by(show_title) |>
  arrange(week) |>
  mutate(
    ever_hit_1 = any(weekly_rank == 1),
    debut_rank = first(weekly_rank),
    latest_week = max(week)
  ) |>
  filter(ever_hit_1 == TRUE, debut_rank > 1) |>
  ungroup() |>
  arrange(desc(latest_week)) |>
  slice(1)
```

# Q10. Which TV show/season hit the top 10 in the most countries in its debut week? In how many countries did it chart?
# Emily in Paris hit the top 10 in 94 countries in its debut week.

```{r}
tv_debuts <- COUNTRY_TOP_10 |>
  filter(str_detect(category, "TV")) |>
  group_by(show_title) |>
  filter(week == min(week)) |>
  summarise(countries_count = n_distinct(country_name), .groups = "drop") |>
  arrange(desc(countries_count)) |>
  slice(1)
```

## Press Releases

