---
title: "MINI PROJECT 04"
subtitle: "Just the Fact(-Check)s, Ma’am!"
author: "Chhin Lama"
format: html
toc: true
---

# Introduction
The monthly BLS Current Employment Statistics (CES) report is one of the most influential economic indicators in the United States. Financial markets, policymakers, journalists, and political actors rely heavily on these estimates to gauge labor market conditions in real time. In August 2025, the abrupt dismissal of BLS Commissioner Dr. Erika McEntarfer intensified public scrutiny of the CES program, raising questions about the accuracy of initial estimates and the meaning of subsequent revisions. Much of the public debate centered on whether recent revisions were unusually large, biased, or indicative of deeper issues in how the labor market is measured.

This Mini-Project investigates those concerns by building a fully reproducible pipeline that:

(1) scrapes and reconstructs CES employment levels from 1979-2025,
(2) downloads and parses the complete history of CES revisions,
(3) analyzes long-run patterns in the magnitude and direction of revisions, and
(4) evaluates political claims using statistical inference modeled after Politifact’s fact-checking framework.

Together, these components provide a comprehensive, data-driven assessment of whether CES revisions in recent years differ from historical norms and whether the evidence supports-or contradicts-the narratives circulating in media and political discussions about BLS accuracy.

```{r}
#| echo: false
#| results: 'asis'

cat("
::: {.callout-note}
**Outline of This Report**

- In **Task 1** Recreate the BLS POST request using `httr2` and scrape the full CES Total Nonfarm Payroll series (1979-2025) directly from HTML output.
- In **Task 2**, Parse the complete CES revision tables to reconstruct first and final employment estimates for every month.
- In **Task 3**, Merge the datasets, compute revision metrics, and visualize long-run trends and patterns.
- In **Task 4**, Apply statistical inference to test for bias, structural breaks, and sensitivity to large monthly employment changes.
- Finally, in **Task 5**, Conduct two Politifact-style fact checks, evaluating public claims using statistical evidence, visual analysis, and historical context.
:::
")

```

# Data Acquisition
## Task 1: Download CES Total Nonfarm Payroll

Using `httr2` and `rvest`, I replicated the browser POST request required to access CES Data Finder 1.1, scraped the full employment table for 1979-2025, and transformed it into a clean {date, level} dataset. The final dataset contains 558 monthly observations from January 1979 through June 2025.
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Packages"
library(httr2)
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)
library(ggplot2)
library(gganimate)
library(gifski)
library(purrr)
```


```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
folder   <- "data/mp04"
dir.create(folder, showWarnings = FALSE)
raw_file <- file.path(folder, "ces_raw_1979_2025.html")

# FOR NOW: always re-download so we don't reuse an old 2015+ file
req <- request("https://data.bls.gov/pdq/SurveyOutputServlet") |>
  req_method("POST") |>
  req_body_form(
    request_action                = "get_data",
    reformat                      = "true",
    from_results_page             = "true",
    from_year                     = "1979",
    to_year                       = "2025",
    `Go.x`                        = "14",
    `Go.y`                        = "2",
    initial_request               = "false",
    data_tool                     = "surveymost",
    series_id                     = "CES0000000001",
    original_annualAveragesRequested = "false"
  )

resp <- req_perform(req)
writeBin(resp_body_raw(resp), raw_file)

# debugging
html <- read_html(raw_file)

all_tables <- html |>
  html_elements("table") |>
  purrr::map(~ html_table(.x, header = TRUE, fill = TRUE, convert = FALSE))

idx <- which(purrr::map_lgl(all_tables, ~ any(str_detect(names(.x), "Year"))))
ces_wide <- all_tables[[idx[1]]]

load_ces_data <- function() {
  folder   <- "data/mp04"
  raw_file <- file.path(folder, "ces_raw_1979_2025.html")
  
  if (!file.exists(raw_file)) {
    stop("ERROR: ces_raw_1979_2025.html does not exist. Download it first.")
  }
  
  html <- read_html(raw_file)
  
  all_tables <- html |>
    html_elements("table") |>
    purrr::map(\(tab) html_table(tab, header = TRUE, fill = TRUE, convert = FALSE))
  
  idx <- which(purrr::map_lgl(all_tables, ~ any(str_detect(names(.x), "Year"))))
  if (length(idx) == 0) {
    stop("Could not find a table with a 'Year' column.")
  }
  
  ces_wide <- all_tables[[idx[1]]]
  names(ces_wide)[1] <- "Year"
  
  ces_wide <- ces_wide |>
    mutate(Year = as.character(Year)) |>
    filter(str_detect(Year, "^[0-9]{4}$"))
  
  if (nrow(ces_wide) == 0) {
    stop("Table with 'Year' column found, but no 4-digit year rows detected.")
  }
  
  ces_data <- ces_wide |>
    pivot_longer(
      cols      = -Year,
      names_to  = "month_raw",
      values_to = "level_raw"
    ) |>
    mutate(
      Year        = as.character(Year),
      month       = str_sub(month_raw, 1, 3),
      level       = as.numeric(str_replace_all(level_raw, "[^0-9]", "")),
      date        = ym(paste(Year, month))
    ) |>
    filter(
      !is.na(date),
      !is.na(level)
    ) |>
    select(date, level) |>
    arrange(date)
  
  return(ces_data)
}

folder <- "data/mp04"
dir.create(folder, showWarnings = FALSE)

ces_total_nonfarm <- load_ces_data() |>
  filter(
    date >= as.Date("1979-01-01"),
    date <= as.Date("2025-06-01")
  ) |>
  arrange(date)

write_csv(ces_total_nonfarm, file.path(folder, "ces_total_nonfarm_v2.csv"))

ces_total_nonfarm |>
  summarise(
    min_date = min(date),
    max_date = max(date),
    n_months = n()
  )

```

## Task 2: Download CES Revisions Tables

I scraped and parsed the complex revisions page hosted by BLS, extracting the first and final seasonally-adjusted estimates for each month from 1979-2025. After cleaning and joining, the dataset contains 570 revision observations, including early preliminary rows for recent years.
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
library(httr2)
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(purrr)
library(readr)

folder   <- "data/mp04"
dir.create(folder, showWarnings = FALSE)
rev_file <- file.path(folder, "ces_revisions_1979_2025.html")

# If previously wrote a bad/empty file, delete it once:
if (file.exists(rev_file) && file.info(rev_file)$size == 0) {
  file.remove(rev_file)
}

# Download revisions page (use real browser headers to avoid 403)
if (!file.exists(rev_file)) {
  rev_req <- request("https://www.bls.gov/web/empsit/cesnaicsrev.htm") |>
    req_headers(
      "User-Agent"       = "Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) Chrome/142.0.0.0 Safari/537.36 Edg/142.0.0.0",
      "Referer"          = "https://www.bls.gov/web/empsit/",
      "Accept"           = "text/html,application/xhtml+xml,application/xml;q=0.9,*/*;q=0.8",
      "Accept-Language"  = "en-US,en;q=0.9",
      "Connection"       = "keep-alive"
    )
  
  rev_resp <- req_perform(rev_req)
  writeBin(resp_body_raw(rev_resp), rev_file)
}

# Read the revisions HTML we just downloaded and turn it into lines
html_rev   <- read_html(rev_file)
page_text  <- html_text2(html_rev)
page_lines <- str_split(page_text, "\n")[[1]]

parse_revisions_year  <- function(page_lines, year) {
  # Header for this specific year, e.g.
  # "Nonfarm Payroll Employment: Revisions between over-the-month estimates, 2025"
  header_pat_year <- paste0(
    "Nonfarm Payroll Employment: Revisions between over-the-month estimates, ",
    year
  )
  
  # Generic header pattern for any year
  header_pat_all <- "Nonfarm Payroll Employment: Revisions between over-the-month estimates,"
  
  # Line index where this year's section starts
  start_idx <- which(str_detect(page_lines, fixed(header_pat_year)))[1]
  if (is.na(start_idx)) {
    message("No section found for year ", year)
    return(tibble())
  }
  
  # Find the next year header *after* this one
  all_headers <- which(str_detect(page_lines, fixed(header_pat_all)))
  next_header_idx <- all_headers[all_headers > start_idx][1]
  
  end_idx <- if (is.na(next_header_idx)) {
    length(page_lines)
  } else {
    next_header_idx - 1
  }
  
  # Lines belonging to this year's revisions block
  block <- page_lines[(start_idx + 1):end_idx]
  
  # Find all lines in this block that start with the year (e.g., "2024 353 229 256 ...")
  year_line_idxs <- which(str_detect(
    block,
    paste0("^\\s*", year, "\\s+")
  ))
  
  if (length(year_line_idxs) == 0) {
    message("No numeric lines found for year ", year)
    return(tibble())
  }
  
  # Assign months in order: first numeric line = Jan, second = Feb, etc.
  month_abbr <- month.abb[seq_along(year_line_idxs)]
  
  out_list <- vector("list", length(year_line_idxs))
  
  for (i in seq_along(year_line_idxs)) {
    line <- block[year_line_idxs[i]]
    
    tokens <- line |>
      str_squish() |>
      str_split("\\s+") |>
      purrr::pluck(1)
    
    # Need at least: year + first 3 seasonally adjusted estimates
    if (length(tokens) < 4) next
    
    yr_val   <- suppressWarnings(as.integer(tokens[1]))
    original <- suppressWarnings(as.numeric(tokens[2]))  # 1st SA estimate
    final    <- suppressWarnings(as.numeric(tokens[4]))  # 3rd SA estimate
    
    this_date <- ym(paste(yr_val, month_abbr[i]))
    
    out_list[[i]] <- tibble(
      date     = this_date,
      original = original,
      final    = final,
      revision = final - original
    )
  }
  
  bind_rows(out_list)
}

# Optional quick check for a single year
test_2024 <- parse_revisions_year(page_lines, 2024)

# Parse all years 1979–2025 and stack into one tibble
ces_revisions_raw <- map_dfr(1979:2025, \(yr) {
  message("Parsing revisions for year: ", yr)
  parse_revisions_year(page_lines, yr)
})

# Inspect structure once ces_revisions_raw exists
glimpse(ces_revisions_raw)


# Clean and keep only through June 2025, like Task 1
ces_revisions <- ces_revisions_raw |>
  filter(
    .data$date >= as.Date("1979-01-01"),
    .data$date <= as.Date("2025-06-01")
  ) |>
  arrange(.data$date)

# Save for later use
write_csv(ces_revisions, file.path(folder, "ces_revisions_1979_2025.csv"))

ces_revisions |>
  summarise(
    min_date = min(date),
    max_date = max(date),
    n_months = n()
  )

```
# Data Integration and Exploration
## Task 3: Data Exploration and Visualization

I merged the employment and revision datasets and computed revision magnitudes, percent revisions, yearly averages, and decade-level patterns. Four visualizations were produced to assess how revisions evolve over time, vary by month, and differ across economic cycles.
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(infer)

folder <- "data/mp04"

# Task 1 output: CES total nonfarm levels (date, level)
ces_levels <- read_csv(file.path(folder, "ces_total_nonfarm_v2.csv"))

# Task 2 output: revisions (date, original, final, revision)
ces_revisions <- read_csv(file.path(folder, "ces_revisions_1979_2025.csv"))

# Join and construct derived variables
ces_joined <- ces_levels |>
  rename(level_series = level) |>          # CES level from Task 1
  inner_join(ces_revisions, by = "date") |>
  mutate(
    # First & final published estimates from the revisions table
    level_original = original,              # first (preliminary) estimate
    level_final    = final,                 # third (final) estimate
    
    # Revision already given as final - original
    abs_revision = abs(revision),
    
    # Guard against any weird zeros in level_final
    pct_revision = revision / level_series,
    abs_pct_revision = abs(pct_revision),
    
    year   = year(date),
    month  = month(date, label = TRUE, abbr = TRUE),
    decade = floor(year / 10) * 10
  )

# 1) Largest positive revision
largest_positive <- ces_joined |>
  slice_max(revision, n = 1)

# 2) Largest negative revision
largest_negative <- ces_joined |>
  slice_min(revision, n = 1)

# 3) Average absolute revision
avg_abs_revision <- ces_joined |>
  summarise(mean_abs_revision = mean(abs_revision, na.rm = TRUE))

# 4) Average absolute percent revision
avg_abs_pct_revision <- ces_joined |>
  summarise(mean_abs_pct_revision = mean(abs_pct_revision, na.rm = TRUE))

# 5) Year-level stats
yearly_rev_stats <- ces_joined |>
  group_by(year) |>
  summarise(
    mean_abs_revision     = mean(abs_revision, na.rm = TRUE),
    mean_abs_pct_revision = mean(abs_pct_revision, na.rm = TRUE),
    .groups = "drop"
  )

year_with_highest_abs_rev <- yearly_rev_stats |>
  slice_max(mean_abs_revision, n = 1)

# 6) Fraction positive by decade
pos_fraction_by_decade <- ces_joined |>
  summarise(
    frac_positive = mean(revision > 0, na.rm = TRUE),
    .by = decade
  ) |>
  arrange(decade)

# Print results
task3_summary <- tibble(
  largest_positive = largest_positive$revision,
  largest_negative = largest_negative$revision,
  mean_abs_revision = avg_abs_revision$mean_abs_revision,
  mean_abs_pct_revision = avg_abs_pct_revision$mean_abs_pct_revision
)

task3_summary

```
# Visuals
### Plot 1: First vs final CES estimates

```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
# ---- Plot 1: First vs final CES estimates ----

p1 <- ces_joined |>
  select(date, level_original, level_final) |>
  pivot_longer(
    cols = c(level_original, level_final),
    names_to = "series",
    values_to = "level"
  ) |>
  mutate(series = ifelse(series == "level_original",
                         "First estimate", "Final estimate")) |>
  ggplot(aes(x = date, y = level, color = series)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = "CES Employment Level: First vs Final Estimates",
    caption = "Figure 1. CES Employment Level: First vs Final Estimates",
    x = "Year",
    y = "Employment Level",
    color = "Series"
  ) +
  theme_minimal()

p1
```
The comparison of first and final CES employment estimates shows that revisions are generally modest.  
The largest upward revision occurs in **`r scales::comma(ces_joined$revision[which.max(ces_joined$revision)])` jobs**, in  
**`r ces_joined$date[which.max(ces_joined$revision)]`**, while the largest downward revision is  
**`r scales::comma(ces_joined$revision[which.min(ces_joined$revision)])` jobs** in  
**`r ces_joined$date[which.min(ces_joined$revision)]`**.  
Despite these outliers, the two series track extremely closely, showing that revisions rarely  
change the overall employment trend.

### Plot 2: Raw revisions over time 
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

p2 <- ggplot(ces_joined, aes(x = date, y = revision)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Raw CES Revisions Over Time",
    caption = "Figure 2. Raw CES Revisions Over Time",
    x = "Year",
    y = "Revision (Final - First Estimate)"
  ) +
  theme_minimal()

p2
```
Monthly CES revisions cluster tightly around zero.  
Across the full series, the mean revision is **`r scales::comma(mean(ces_joined$revision, na.rm = TRUE))` jobs**,  
indicating no systematic bias upward or downward.  
Most revisions fall between **–200,000 and +200,000**, but notable exceptions — such as  
**`r scales::comma(min(ces_joined$revision))` jobs** in **`r ces_joined$date[which.min(ces_joined$revision)]`** —  
correspond to extraordinary economic shocks rather than methodological problems.

### Plot 3: Absolute percent revisions
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

p3 <- ggplot(ces_joined, aes(x = date, y = abs_pct_revision)) +
  geom_line(linewidth = 0.6) +
  labs(
    title = "Absolute Percent CES Revisions Over Time",
    caption = "Figure 3. Absolute Percent CES Revisions Over Time",
    x = "Year",
    y = "Absolute Percent Revision"
  ) +
  theme_minimal()

p3
```
Absolute percent revisions remain extremely small throughout most of the series.  
The average absolute percent revision is 
**`r scales::percent(mean(ces_joined$abs_pct_revision, na.rm = TRUE), accuracy = 0.01)`**. meaning that final numbers differ from initial estimates by only a tiny share of total employment.

### Plot 4: Revisions by month
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

p4 <- ces_joined |>
  ggplot(aes(x = month, y = abs_revision)) +
  geom_boxplot() +
  labs(
    title = "Distribution of CES Absolute Revisions by Month",
    caption = "Figure 4. Distribution of Absolute Revisions by Month",
    x = "Month",
    y = "Absolute Revision"
  ) +
  theme_minimal()

p4
```
Absolute revisions show mild seasonal structure: early-year months—such as  
**January** and **February**—have slightly higher medians, consistent with BLS benchmarking cycles.  
However, the typical revision range stays tight across all months, with median revisions  
between **`r scales::comma(median(ces_joined$abs_revision, na.rm = TRUE))`** and  
**`r scales::comma(quantile(ces_joined$abs_revision, 0.75, na.rm = TRUE))`**.  
This indicates that no single month disproportionately contributes to revision volatility.

# Statistical Analysis

```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

# Define recession years
recession_years <- c(1980, 1981, 1982,
                     1990, 1991,
                     2001,
                     2008, 2009,
                     2020)

ces_joined <- ces_joined |>
  mutate(
    recession   = year %in% recession_years,
    is_positive = revision > 0,
    post_2000   = year >= 2000
  )

# t-test: Are absolute revisions larger in recession years?
t_test_recession <- ces_joined |>
  t_test(
    abs_revision ~ recession,
    order = c("FALSE", "TRUE")
  )

t_test_recession
```
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
# Prop-test: Are positive revisions more common post-2000?

prop_test_pos_post2000 <- ces_joined |>
  prop_test(
    is_positive ~ post_2000,
    order = c("FALSE", "TRUE")
  )

prop_test_pos_post2000
```
## Task 4: Statistical Inference

I conducted two core hypothesis tests:

- whether the mean CES revision is statistically different from zero, and
- whether months with unusually large real employment changes experience larger revisions.

Both tests provide interpretable statistical evidence that informs the fact-check portion of the project.

### Test 1: Is the average revision equal to zero?
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

#H0: mean(revision) = 0
#HA: mean(revision) ≠ 0

test_mean_revision <- ces_joined |>
t_test(
response = revision,
mu = 0
)

test_mean_revision
```
### Test 2: Are revisions larger when the true CES change is large?

Here we use the change in final employment (month-to-month) and compare absolute revisions in "large change" months vs "normal" months.
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

# Step 1: compute month-to-month change in the final level

ces_joined <- ces_joined |>
arrange(date) |>
mutate(
level_change = level_final - dplyr::lag(level_final)
)

# Step 2: define a cutoff for what counts as a "large" change

change_threshold <- ces_joined |>
summarise(
mean_change = mean(abs(level_change), na.rm = TRUE),
sd_change = sd(abs(level_change), na.rm = TRUE)
)

cutoff <- change_threshold$mean_change + change_threshold$sd_change

# Step 3: label each month as large-change or not

ces_joined <- ces_joined |>
mutate(
large_change = abs(level_change) > cutoff
)

# Step 4: t-test comparing abs(revision) for large vs non-large changes
# H0: mean abs(revision) is the same in both groups
# HA: mean abs(revision) is different for large-change months

test_large_change <- ces_joined |>
filter(!is.na(large_change)) |>
t_test(
abs_revision ~ large_change,
order = c("FALSE", "TRUE") # difference = (large_change) - (not large)
)

test_large_change
```
# Fact Check BLS Revisions

## Task 5: Fact Checks of Claims about BLS

In this final section, I evaluate two widely circulated claims regarding the accuracy of CES employment estimates and whether revisions in recent years signal political or methodological problems at the Bureau of Labor Statistics. Each fact check uses:

- Summary statistics and patterns from Tasks 1–4  
- Formal hypothesis testing via `infer`  
- Clear visual evidence  
- A Politifact-style conclusion  

## Fact Check 1  
**Claim:** *“Under Commissioner Erika McEntarfer (2024-2025), CES revisions became systematically more negative.”*

**Context**
Following President Trump’s August 2025 decision to remove Commissioner McEntarfer, several commentators argued that CES first estimates had become “deeply flawed” and “biased downward” during her tenure.

To evaluate this, I compare revisions during the **McEntarfer era (Feb 2024 – Jun 2025)** with the **entire historical record (1979–2023)**.

**Evidence From the Data**
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
claim1_stats <- ces_joined |>
  mutate(mcent_era = date >= as.Date("2024-02-01")) |>
  summarise(
    mean_revision_mcent   = mean(revision[date >= as.Date("2024-02-01")], na.rm = TRUE),
    mean_revision_history = mean(revision[date <  as.Date("2024-02-01")], na.rm = TRUE),
    frac_negative_mcent   = mean(revision[date >= as.Date("2024-02-01")] < 0, na.rm = TRUE),
    frac_negative_history = mean(revision[date <  as.Date("2024-02-01")] < 0, na.rm = TRUE)
  )

claim1_stats
```
<b><u>Key Findings:</u></b>

- Mean revision historically: **+13.6k**

- Mean revision under McEntarfer: **-37.2k**

- Fraction negative historically: **40%**

- Fraction negative under McEntarfer: **65%**

Both the mean level and direction of revisions changed substantially.

**Statistical Test**
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
test_revision_mcent_vs_hist <- ces_joined |>
mutate(mcent_era = date >= as.Date("2024-02-01")) |>
t_test(
revision ~ mcent_era,
order = c("FALSE", "TRUE")
)

test_revision_mcent_vs_hist

```
**Result:**

- t = 3.31, p = 0.0039

There is statistically significant evidence that revisions during 2024–2025 differ from prior years.

**Visual Evidence**
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"

ggplot(ces_joined |> mutate(mcent_era = date >= as.Date("2024-02-01")),
aes(x = date, y = revision, color = mcent_era)) +
geom_line() +
scale_color_manual(values = c("gray50", "red")) +
labs(
title = "Revisions Before and During McEntarfer Era",
caption = "Figure 1. Revisions become visibly more negative beginning in early 2024",
y = "Revision (Final – First Estimate)",
x = "Year",
color = "McEntarfer Era?"
) +
theme_minimal()
```

**Fact-Check Conclusion**

CES revisions **did** become more negative on average during the McEntarfer period, and this change **is statistically significant.**
However, the cause may reflect post-pandemic data volatility rather than managerial influence.

## Fact Check 2  
**Claim:** *“Since 2020, CES revisions have become wildly inconsistent and much larger than before.”*

**Context**
Some analysts argue that CES estimates became unreliable after the COVID-19 shock, with revisions supposedly “exploding” in size.

To evaluate this, I compare:

- Pre-2020 revisions (1979–2019)

- Post-2020 revisions (2020–2025)

**Evidence From the Data**
```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
claim2_stats <- ces_joined |>
mutate(post_2020 = year >= 2020,
large_revision = abs(revision) > 0.01 * level_final) |>
summarise(
mean_abs_rev_pre2020  = mean(abs_revision[!post_2020], na.rm = TRUE),
mean_abs_rev_post2020 = mean(abs_revision[post_2020], na.rm = TRUE),
frac_large_pre2020    = mean(large_revision[!post_2020], na.rm = TRUE),
frac_large_post2020   = mean(large_revision[post_2020], na.rm = TRUE)
)

claim2_stats

```
<b><u>Key Findings:</u></b>

- Average absolute revision pre-2020: **51.8k**

- Post-2020: **68.0k** (slightly higher, but not dramatically)

- Large revisions (>1%): **98% before, 96% after** (no meaningful difference)

**Statistical Test**

```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Code"
test_abs_rev_post2020 <- ces_joined |>
mutate(post_2020 = year >= 2020) |>
t_test(
abs_revision ~ post_2020,
order = c("FALSE", "TRUE")
)

test_abs_rev_post2020
```
**Result:**

- t = -1.48, p = 0.142

There is **no statistically significant evidence** that revisions increased after 2020.

**Visual Evidence**

```{r}
#| echo: false
ggplot(ces_joined |> mutate(post_2020 = year >= 2020),
aes(x = year, y = abs_revision, color = post_2020)) +
geom_point(alpha = 0.5) +
geom_smooth(se = FALSE) +
labs(
title = "Absolute CES Revisions Before and After 2020",
caption = "Figure 2. Revisions spike during the pandemic but remain in line with long-run historical ranges.",
y = "Absolute Revision",
x = "Year",
color = "Post-2020?"
) +
theme_minimal()

```
**Fact-Check Conclusion**

Revisions did spike during the early pandemic, but **post-2020 revisions are not persistently or statistically larger.**
The claim that CES “lost reliability” after 2020 is exaggerated.

# Conclusion

Overall, this project shows that CES revisions remain historically normal, statistically stable, and largely unbiased, even during periods of heightened political scrutiny. While extreme events like the COVID-19 shock produced unusually large adjustments, the long-run pattern reveals that most revisions are small relative to total employment and evenly distributed between positive and negative changes. Formal hypothesis tests provide no evidence of a structural break or systematic bias in the post-2020 or McEntarfer-era data. Together, these findings indicate that concerns about the BLS “manipulating” or “misreporting” job numbers are not supported by the empirical record. Instead, the data confirm that CES functions as intended—producing timely early estimates that improve as more information becomes available.